A formalization of semi-simplicial and semi-cubical sets

### 2023-01-09 paper/paper.tex
paper: start preparing for LICS submission
The last table in landscape mode isn't properly formatted.

\usepackage{multirow, float, tabularx, environ, booktabs, setspace, makecell, caption, art.cls/joinargs, colortbl, pdflscape, minted}
\usepackage[prefix=bonak]{art.cls/xkeymask}
\usepackage[skip=0.4em]{parskip}
\usetikzlibrary{patterns}

@@

% The eqntable environment, displaying the various
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\NewEnviron{eqntable}[1]{
  \begin{table}[H]
    \small
    \begin{tabularx}{\linewidth}{@{}>{$}l<{$} >{$}c<{$} >{$}c<{$} >{$}Y<{$} @{}}
      \toprule
      \BODY
      \bottomrule
    \end{tabularx}
    \caption{#1}
  \end{table}
}

@@

\title{A formalization of semi-simplicial and semi-cubical sets}
\author{
  \textcolor{gray80}{Hugo Herbelin} \\
  \itshape \textcolor{gray80}{Université Paris Cité, Inria, CNRS, IRIF, Paris}
  \and
  \textcolor{gray80}{Ramkumar Ramachandra} \\
  \itshape \textcolor{gray80}{Université Paris Cité, Inria, CNRS, IRIF, Paris}
}
\date{}

@@

\begin{document}
\maketitle
\begin{abstract}
  We present a parametricity-based construction of semi-simplicial sets and semi-cubical sets in indexed form, formulated in dependent type theory. The accompanying artifact is an axiom-less~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to $\nu$-types; if $\nu$ is finite (say, unit type corresponding to augmented semi-simplicial types), this requirement vanishes.} Coq formalization that uses sophisticated rewriting techniques and pushes the boundaries of proof assistant technology.
\end{abstract}
\tableofcontents
\newpage

@@

\begin{definition}[$\DeltaPlus$]
  \begin{align*}
    \obj(\DeltaPlus) & := \Nat                                                                     \\
    \hom(p, n)       & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in l = p}\}        \\
    g \circ f        & :=
    \begin{cases}
      f                  & \text{if $g = \{\}$}                                                      \\
      0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
      a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
    \end{cases} \\
    \id              & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
  \end{align*}

  Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.
\end{definition}

@@

\subsection{Semi-cubical sets}
\begin{definition}[\Cube]
  $\Cube$ is defined similarly to $\DeltaPlus$, except that we take sequences of $L$, $R$ and $\star$, instead of sequences of $0$ and $\kstar$. We use $L$ and $R$ to indicate opposite faces of a cube.

  \begin{align*}
    \obj(\Cube) & := \Nat                                                                                \\
    \hom(p, n)  & := \{l \in \{L, R, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
    g \circ f   & :=
    \begin{cases}
      f                  & \text{if $g = \{\}$}                                                            \\
      a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
      a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
    \end{cases} \\
    \id         & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
  \end{align*}

@@

\begin{align*}
  X_0 & : \U                                                                                                              \\
  X_1 & : X_0 \times X_0 \rightarrow \U                                                                                   \\
  X_2 & : \forall a b c d, X_1 (a, b) \rightarrow X_1 (a, c) \rightarrow X_1 (b, d) \rightarrow X_1 (c, d) \rightarrow \U \\
  \ldots
\end{align*}

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a : A, (B a \rightarrow C)$ and $(\Sigma a : A, B a) \rightarrow C$. Let us illustrate how we build $\fullframe[n]$ by refining types of $X_0, X_1, \ldots, X_{n - 1}$ so that they have the form $\fullframe[n] \rightarrow \U$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Each $\fullframe[n]$ is made of $n$-layers, written $\layer[n][p]$ with $p < n$, initialized with the $\unittype$ type. Then, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$. The rest of the construction follows.

\begin{equation*}
  \begin{array}{ll}
    X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
    X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
    \begin{array}{c}
        \underbrace{X_0(\unitpoint)}_{\filler[0][0]}
        \\ \times \\
        \underbrace{X_0(\unitpoint)}_{\filler[0][0]}
      \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
    X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
        \begin{array}{c}
          \underbrace{\Sigma b: \left(
          \hspace{-0.4em}\begin{array}{c}
                           X_0(\unitpoint)
                           \\ \times \\
                           X_0(\unitpoint)
                         \end{array}\hspace{-0.4em}
          \right).\; \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L]}}_{\filler[1][1]}}_{\filler[1][0]}
          \\ \times \\
          \underbrace{\Sigma b: \left(
          \hspace{-0.4em}\begin{array}{c}
                           X_0(\unitpoint)
                           \\ \times \\
                           X_0(\unitpoint)
                         \end{array}\hspace{-0.4em}
          \right).\; \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R]}}_{\filler[1][1]}}_{\filler[1][0]}
        \end{array}
        \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
      \begin{array}{c}
        \underbrace{X_1 \underbrace{\left(a.\hd, \left(
            \hspace{-0.4em}\begin{array}{c}
                             a.\tl.L.\hd.L, \\
                             a.\tl.R.\hd.L
                           \end{array}\hspace{-0.4em}
            \right)\right)}_{\restrf[2][1][L]}}_{\filler[1][1]}
        \\ \times \\
        \underbrace{X_1 \underbrace{\left(a.\hd, \left(
            \hspace{-0.4em}\begin{array}{c}
                             a.\tl.L.\hd.R, \\
                             a.\tl.R.\hd.R
                           \end{array}\hspace{-0.4em}
            \right)\right)}_{\restrf[2][1][R]}}_{\filler[1][1]}
      \end{array}
      \right)}_{\layer[2][1]}}_{\framep[2][2]}
    \rightarrow \U                                                                                                                                \\
    \ldots
  \end{array}
\end{equation*}

@@

\begin{figure}[H]
  \centering
  \begin{tikzpicture}[scale=2]
    \draw[spanish-blue, fill=spanish-blue] (0, 0) -- (1, 0) -- (1, 1) -- (0, 1) -- (0, 0);
    \draw[spanish-blue, fill=spanish-blue, nearly transparent] (0.6, 1) -- (0.6, 1.6) -- (1.6, 1.6) -- (1.6, 0.6) -- (1, 0.6);
  \end{tikzpicture}
  \;\;
  \begin{tikzpicture}[scale=2]
    \draw[spanish-blue, fill=spanish-blue] (0, 0) -- (1, 0) -- (1, 1) -- (0, 1) -- (0, 0);
    \draw[spanish-blue, fill=spanish-blue, nearly transparent] (0.6, 1) -- (0.6, 1.6) -- (1.6, 1.6) -- (1.6, 0.6) -- (1.0, 0.6);
    \draw[raspeberry, pattern=dots] (1.1, 1.1) -- (1.5, 1.5) -- (1.5, 0.5) -- (1.1, 0.1) -- (1.1, 1.1);
    \draw[raspeberry] (1.3, 1.3) -- (1.3, 0.3);
    \draw[raspeberry, pattern=dots] (0.1, 1.1) -- (0.5, 1.5) -- (0.5, 1.1);
    \draw[raspeberry] (0.3, 1.3) -- (0.3, 1.1);
  \end{tikzpicture}
  \;\;
  \begin{tikzpicture}[scale=2]
    \draw[spanish-blue, fill=spanish-blue] (0, 0) -- (1, 0) -- (1, 1) -- (0, 1) -- (0, 0);
    \draw[spanish-blue, fill=spanish-blue, nearly transparent] (0.6, 1) -- (0.6, 1.6) -- (1.6, 1.6) -- (1.6, 0.6) -- (1.0, 0.6);
    \draw[raspeberry, pattern=dots] (1.1, 1.1) -- (1.5, 1.5) -- (1.5, 0.5) -- (1.1, 0.1) -- (1.1, 1.1);
    \draw[raspeberry] (1.3, 1.3) -- (1.3, 0.3);
    \draw[raspeberry] (0.1, 1.1) -- (0.5, 1.5);
    \draw[russian-green, pattern=dots] (0.2, 1.1) -- (1.0, 1.1) -- (1.4, 1.5) -- (0.6, 1.5) -- (0.2, 1.1);
    \filldraw[russian-green] (0.8, 1.3) circle (0.6pt);
  \end{tikzpicture}
\end{figure}

@@

\begin{align*}
  \framep[n][p](D) & \defeq \Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0](D)) \ldots) .\; \layer[n][p](D) \\
  \filler[n][p](D) & \defeq \Sigma l_p : \layer[n][p](D) . (\ldots (\Sigma l_n : \layer[n][n - 1](D) .\; D.\tl) \ldots)
\end{align*}

@@

\begin{align*}
  \framep[n][p](D)    & \defeq \Sigma d : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0](D)(\unitpoint)) \ldots) .\; \layer[n][p](D)(d)                              \\
  \filler[n][p](D)(d) & \defeq \Sigma l_p : \layer[n][p](D)(d) . (\ldots (\Sigma l_n : \layer[n][n - 1](D)(d, l_p, \ldots, l_{n - 1}) .\; D.\tl(d, l_p, \ldots, l_n)) \ldots) \\
  \text{where}\;      & (d, l_p, \ldots, l_q) \text{ abbreviates } ((\ldots(d, l_p), \ldots), l_q)
\end{align*}

@@

\begin{align*}
  \layer[n][p](D)(d) & \defeq \filler[n-1][p](D.\hd)(\restrf[n][p][L][p](d)) \times \filler[n-1][p](D.\hd)(\restrf[n][p][R][p](d))
\end{align*}

@@

\begin{landscape}
  \renewcommand*{\lab}{fulltab:faces}
  \input{tab-faces.tex}

  \renewcommand*{\lab}{fulltab:coh}
  \input{tab-coh.tex}
\end{landscape}

>>>>

\usepackage{multirow, float, tabularx, environ, booktabs, setspace, makecell, caption, art.cls/joinargs, colortbl, multicol, pdflscape, minted}
\usepackage[prefix=bonak]{art.cls/xkeymask}
\usepackage[skip=0.4em]{parskip}
\usetikzlibrary{patterns}

@@

% The eqntable environment, displaying the various
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\NewEnviron{onecoltbl}[1]{
  \begin{table}[H]
    \small
    \begin{tabularx}{\linewidth}{@{}>{$}l<{$} >{$}c<{$} >{$}c<{$} >{$}Y<{$} @{}}
      \toprule
      \BODY
      \bottomrule
    \end{tabularx}
    \caption{#1}
  \end{table}
}

\NewEnviron{eqntable}[1]{
  \begin{table*}[t!]
    \small
    \begin{tabularx}{\linewidth}{@{}>{$}l<{$} >{$}c<{$} >{$}c<{$} >{$}Y<{$} @{}}
      \toprule
      \BODY
      \bottomrule
    \end{tabularx}
    \caption{#1}
  \end{table*}
}

@@

\title{A formalization of semi-simplicial and semi-cubical sets}
\author{
  Hugo Herbelin \\
  Université Paris Cité, Inria, CNRS, IRIF, Paris \\
  \and
  Ramkumar Ramachandra \\
  Université Paris Cité, Inria, CNRS, IRIF, Paris
}
\date{}

@@

\begin{document}
\maketitle
\begin{abstract}
  We present a parametricity-based construction of semi-simplicial sets and semi-cubical sets in indexed form, formulated in dependent type theory. The accompanying artifact is an axiom-less~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to $\nu$-types; if $\nu$ is finite (say, unit type corresponding to augmented semi-simplicial types), this requirement vanishes.} Coq formalization that uses sophisticated rewriting techniques and pushes the boundaries of proof assistant technology.
\end{abstract}

@@

\begin{definition}[$\DeltaPlus$]
  \begin{figure*}[!t]
    \begin{align*}
      \obj(\DeltaPlus) & := \Nat                                                                     \\
      \hom(p, n)       & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in l = p}\}        \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                      \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
    \end{align*}
    \caption{Definition of \DeltaPlus}\label{fig:deltaplus}
  \end{figure*}

  The definition of $\DeltaPlus$ is shown in Fig. \ref{fig:deltaplus}. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.
\end{definition}

@@

\subsection{Semi-cubical sets}
\begin{definition}[\Cube]
  $\Cube$ is defined similarly to $\DeltaPlus$, except that we take sequences of $L$, $R$ and $\star$, instead of sequences of $0$ and $\kstar$. We use $L$ and $R$ to indicate opposite faces of a cube. The definition is shown in Fig. \ref{fig:cube}.

  \begin{figure*}[!t]
    \begin{align*}
      \obj(\Cube) & := \Nat                                                                                \\
      \hom(p, n)  & := \{l \in \{L, R, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = \{\}$}                                                            \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := \{\kstar, \ldots, \kstar\} \text{ $n$ times}
    \end{align*}
    \caption{Definition of \Cube}\label{fig:cube}
  \end{figure*}

@@

\begin{align*}
  X_0 & : \U                                                                        \\
  X_1 & : X_0 \times X_0 \rightarrow \U                                             \\
  X_2 & : \forall a b c d, X_1 (a, b) \rightarrow X_1 (a, c) \rightarrow X_1 (b, d) \\
      & \rightarrow X_1 (c, d) \rightarrow \U                                       \\
  \ldots
\end{align*}

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a : A, (B a \rightarrow C)$ and $(\Sigma a : A, B a) \rightarrow C$. Let us illustrate how we build $\fullframe[n]$ by refining types of $X_0, X_1, \ldots, X_{n - 1}$ so that they have the form $\fullframe[n] \rightarrow \U$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Each $\fullframe[n]$ is made of $n$-layers, written $\layer[n][p]$ with $p < n$, initialized with the $\unittype$ type. Then, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$. The rest of the construction is shown in Fig. \ref{fig:intuition}

\begin{figure*}[!t]
  \begin{equation*}
    \begin{array}{ll}
      X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
      X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
      \begin{array}{c}
          \underbrace{X_0(\unitpoint)}_{\filler[0][0]}
          \\ \times \\
          \underbrace{X_0(\unitpoint)}_{\filler[0][0]}
        \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
      X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
          \begin{array}{c}
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\; \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L]}}_{\filler[1][1]}}_{\filler[1][0]}
            \\ \times \\
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\; \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R]}}_{\filler[1][1]}}_{\filler[1][0]}
          \end{array}
          \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
        \begin{array}{c}
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.L, \\
                               a.\tl.R.\hd.L
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][L]}}_{\filler[1][1]}
          \\ \times \\
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.R, \\
                               a.\tl.R.\hd.R
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][R]}}_{\filler[1][1]}
        \end{array}
        \right)}_{\layer[2][1]}}_{\framep[2][2]}
      \rightarrow \U                                                                                                                                \\
      \ldots
    \end{array}
  \end{equation*}
  \caption{Intuition for formal construction}\label{fig:intuition}
\end{figure*}

@@

\begin{figure}[H]
  \centering
  \begin{tikzpicture}[scale=2]
    \draw[spanish-blue, fill=spanish-blue] (0, 0) -- (1, 0) -- (1, 1) -- (0, 1) -- (0, 0);
    \draw[spanish-blue, fill=spanish-blue, nearly transparent] (0.6, 1) -- (0.6, 1.6) -- (1.6, 1.6) -- (1.6, 0.6) -- (1, 0.6);
  \end{tikzpicture}
  \;\;
  \begin{tikzpicture}[scale=2]
    \draw[spanish-blue, fill=spanish-blue] (0, 0) -- (1, 0) -- (1, 1) -- (0, 1) -- (0, 0);
    \draw[spanish-blue, fill=spanish-blue, nearly transparent] (0.6, 1) -- (0.6, 1.6) -- (1.6, 1.6) -- (1.6, 0.6) -- (1.0, 0.6);
    \draw[raspeberry, pattern=dots] (1.1, 1.1) -- (1.5, 1.5) -- (1.5, 0.5) -- (1.1, 0.1) -- (1.1, 1.1);
    \draw[raspeberry] (1.3, 1.3) -- (1.3, 0.3);
    \draw[raspeberry, pattern=dots] (0.1, 1.1) -- (0.5, 1.5) -- (0.5, 1.1);
    \draw[raspeberry] (0.3, 1.3) -- (0.3, 1.1);
  \end{tikzpicture}
  \newline\newline
  \begin{tikzpicture}[scale=2]
    \draw[spanish-blue, fill=spanish-blue] (0, 0) -- (1, 0) -- (1, 1) -- (0, 1) -- (0, 0);
    \draw[spanish-blue, fill=spanish-blue, nearly transparent] (0.6, 1) -- (0.6, 1.6) -- (1.6, 1.6) -- (1.6, 0.6) -- (1.0, 0.6);
    \draw[raspeberry, pattern=dots] (1.1, 1.1) -- (1.5, 1.5) -- (1.5, 0.5) -- (1.1, 0.1) -- (1.1, 1.1);
    \draw[raspeberry] (1.3, 1.3) -- (1.3, 0.3);
    \draw[raspeberry] (0.1, 1.1) -- (0.5, 1.5);
    \draw[russian-green, pattern=dots] (0.2, 1.1) -- (1.0, 1.1) -- (1.4, 1.5) -- (0.6, 1.5) -- (0.2, 1.1);
    \filldraw[russian-green] (0.8, 1.3) circle (0.6pt);
  \end{tikzpicture}
\end{figure}

@@

\begin{align*}
  \framep[n][p](D) & \defeq & \Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .        \\
                   &        & \layer[n][0](D)) \ldots) .\; \layer[n][p](D)                 \\
  \filler[n][p](D) & \defeq & \Sigma l_p : \layer[n][p](D) .                               \\
                   &        & (\ldots (\Sigma l_n : \layer[n][n - 1](D) .\; D.\tl) \ldots)
\end{align*}

@@

\begin{align*}
  \framep[n][p](D)    & \defeq & \Sigma d : (\ldots (\Sigma \unitpoint: \unittype .\; \layer[n][0](D)(\unitpoint)) \ldots) . \\
                      &        & \layer[n][p](D)(d)                                                                          \\
  \filler[n][p](D)(d) & \defeq & \Sigma l_p: \layer[n][p](D)(d) .                                                            \\
                      &        & (\ldots (\Sigma l_n : \layer[n][n - 1](D)(d, l_p, \ldots, l_{n - 1}) .                      \\
                      &        & D.\tl(d, l_p, \ldots, l_n)) \ldots)                                                         \\
  \text{where}\;      &        & (d, l_p, \ldots, l_q) \text{ abbreviates }                                                  \\
                      &        & ((\ldots(d, l_p), \ldots), l_q)
\end{align*}

@@

\begin{align*}
  \layer[n][p](D)(d) & \defeq & \filler[n-1][p](D.\hd)(\restrf[n][p][L][p](d)) \times \\
                     &        & \filler[n-1][p](D.\hd)(\restrf[n][p][R][p](d))
\end{align*}

@@

\renewcommand*{\lab}{fulltab:faces}
\input{tab-faces.tex}

\begin{landscape}
  \renewcommand*{\lab}{fulltab:coh}
  \input{tab-coh.tex}
\end{landscape}
